---
title: "Combining Genomic Relationship Matrices with Factor-Analytic Models"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    keep_tex: false
    latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Combining Genomic Relationship Matrices with Factor-Analytic Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
```

# Introduction

## Motivation

In plant and animal breeding, genomic relationship matrices (GRMs) quantify
genetic similarity between individuals based on molecular marker data.  Modern
breeding programmes routinely genotype cohorts on **different SNP platforms**
(e.g., low-density vs. high-density chips, or successive generations of arrays
with partially overlapping marker panels).  Each platform produces a GRM over a
potentially different subset of individuals, and the resulting matrices share
some individuals but differ in which markers contribute to the relatedness
estimates.

Combining these partial GRMs into a single, coherent relationship matrix is
essential for:

- **Multi-environment genomic prediction** across cohorts genotyped on different
  platforms,
- **Meta-analysis** of breeding values from independent trials, and
- **Imputation-free workflows** where raw genotypes are unavailable but GRMs
  from each platform are.

## Why a Factor Model?

When partial GRMs cover overlapping but non-identical sets of individuals, the
combined matrix has entries (individual pairs) that are **never jointly observed**
in any single GRM.  The unconstrained (free-$\Sigma$) Wishart EM model cannot
identify those entries --- different starting values give different answers.

A **factor model** ($\Sigma = \Lambda\Lambda^\top + \Psi$) resolves this: if
individual $i$'s factor loading $\Lambda[i,\cdot]$ is determined by their
co-occurrence with shared individuals in one study, and individual $j$'s loading
is determined in another study, then
$\sigma_{ij} = \Lambda[i,\cdot] \cdot \Lambda[j,\cdot]$
becomes identified through shared anchors --- even if $i$ and $j$ never appear
in the same GRM.

This is biologically natural: GRMs are driven by a small number of principal
axes of population structure, and a low-rank factor model captures this
parsimoniously.

## Data: CIMMYT Wheat Lines (BGLR Package)

We use the wheat dataset from the `BGLR` package, which contains:

- **599 CIMMYT wheat lines** genotyped at **1,279 DArT markers**
- **Grain yield** measured in **4 environments**

This is a classic benchmark for genomic selection methods.  We simulate a
realistic multi-platform genotyping scenario by partitioning the markers into
overlapping "SNP chips" and the lines into partially overlapping cohorts, so
that **many individual pairs are never jointly observed**.

# Setup and Data Preparation

```{r load-packages}
library(CovCombR)
library(BGLR)
library(ggplot2)
library(reshape2)

data(wheat)
X <- wheat.X   # 599 x 1279 marker matrix
Y <- wheat.Y   # 599 x 4 phenotype matrix
```

## Full-Data GRM (Ground Truth)

We compute the GRM from the complete marker matrix as the ground truth.

```{r full-grm}
compute_grm <- function(M) {
  Ms <- scale(M, center = TRUE, scale = TRUE)
  Ms[is.nan(Ms)] <- 0
  tcrossprod(Ms) / ncol(Ms)
}

G_full <- compute_grm(X)
n_total <- nrow(G_full)
line_ids <- paste0("L", seq_len(n_total))
dimnames(G_full) <- list(line_ids, line_ids)

cat(sprintf("Full GRM: %d x %d\n", n_total, n_total))
cat(sprintf("Diagonal range: [%.3f, %.3f]\n",
            min(diag(G_full)), max(diag(G_full))))
```

## Multi-Platform Design with Unobserved Pairs

We select 40 wheat lines and split them across three cohorts genotyped on
different marker platforms.  The cohorts use a **chain design** so that:

- Each cohort covers about 15--20 lines,
- Adjacent cohorts share ~5 "anchor" lines,
- Many individual pairs appear in **no** cohort together (e.g., a line only in
  Cohort 1 and a line only in Cohort 3 are never jointly observed).

```{r define-design}
set.seed(2026)

# Select 40 lines for the demo
demo_line_idx <- sort(sample(1:n_total, 40))
demo_ids <- line_ids[demo_line_idx]

# Chain design: C1 and C2 share anchors, C2 and C3 share anchors,
# but C1 and C3 share only a few lines (through wrap-around).
cohort_idx <- list(
  Cohort1 = demo_ids[1:20],               # lines 1--20
  Cohort2 = demo_ids[16:35],              # lines 16--35 (overlap with C1: 16--20)
  Cohort3 = demo_ids[c(31:40, 1:5)]       # lines 31--40 + 1--5 (overlap with C2: 31--35, C1: 1--5)
)

# Different marker platforms
platform_markers <- list(
  ChipA = 1:800,                           # 800 markers
  ChipB = 401:1279,                        # 879 markers
  ChipC = c(1:500, 901:1279)               # 879 markers
)

cat("=== Cohort sizes ===\n")
for (nm in names(cohort_idx))
  cat(sprintf("  %s: %d lines\n", nm, length(cohort_idx[[nm]])))

# Build the joint-observation matrix to identify observed/unobserved pairs
n_demo <- length(demo_ids)
observed_mat <- matrix(FALSE, n_demo, n_demo,
                       dimnames = list(demo_ids, demo_ids))
for (nm in names(cohort_idx)) {
  ids <- cohort_idx[[nm]]
  observed_mat[ids, ids] <- TRUE
}

n_pairs <- sum(upper.tri(observed_mat))
n_obs   <- sum(observed_mat[upper.tri(observed_mat)])
n_unobs <- n_pairs - n_obs

cat(sprintf("\n=== Pair coverage ===\n"))
cat(sprintf("  Total unique pairs: %d\n", n_pairs))
cat(sprintf("  Observed (in >= 1 GRM): %d (%.1f%%)\n",
            n_obs, 100 * n_obs / n_pairs))
cat(sprintf("  UNOBSERVED (never jointly seen): %d (%.1f%%)\n",
            n_unobs, 100 * n_unobs / n_pairs))
```

## Building the Partial GRMs

Each cohort is genotyped on its assigned platform.

```{r build-partial-grms}
platforms_map <- list(Cohort1 = "ChipA", Cohort2 = "ChipB", Cohort3 = "ChipC")

S_list <- list()
nu_vec <- c()

for (nm in names(cohort_idx)) {
  ids <- cohort_idx[[nm]]
  # Map demo IDs back to row indices in the full marker matrix
  row_idx <- demo_line_idx[match(ids, demo_ids)]
  chip <- platforms_map[[nm]]
  marker_idx <- platform_markers[[chip]]

  G <- compute_grm(X[row_idx, marker_idx, drop = FALSE])
  dimnames(G) <- list(ids, ids)
  S_list[[nm]] <- G
  nu_vec[nm] <- length(marker_idx) - 1

  cat(sprintf("  %s: %d x %d GRM from %s (%d markers, nu = %d)\n",
              nm, nrow(G), ncol(G), chip, length(marker_idx), nu_vec[nm]))
}
```

# Combining GRMs with CovCombR

## Choosing the Number of Factors

For GRMs, `n_factors = "auto"` can be slow because the BIC search grid grows
with matrix size.  Since population structure in wheat is well-captured by a few
principal axes, we directly fit factor models with $k = 2, 3, 5$ and compare.
This is the recommended approach for large GRMs.

```{r fit-models}
fit_2f   <- fit_covcomb(S_list, nu_vec, n_factors = 2, se_method = "none")
fit_3f   <- fit_covcomb(S_list, nu_vec, n_factors = 3, se_method = "none")
fit_5f   <- fit_covcomb(S_list, nu_vec, n_factors = 5, se_method = "none")
fit_free <- fit_covcomb(S_list, nu_vec, n_factors = NULL, se_method = "none")

cat("=== Convergence ===\n")
for (info in list(
  list("2-factor", fit_2f), list("3-factor", fit_3f),
  list("5-factor", fit_5f), list("Free",     fit_free)
)) {
  f <- info[[2]]
  cat(sprintf("  %-10s %3d iterations, converged = %s\n",
              info[[1]], f$convergence$iterations, f$convergence$converged))
}
```

## Model Comparison via BIC

```{r model-bic}
bic_row <- function(fit, label) {
  ll <- tail(fit$history$log_likelihood, 1)
  p  <- nrow(fit$Sigma_hat)
  n_nu <- sum(nu_vec)
  if (!is.null(fit$n_factors) && !is.na(fit$n_factors)) {
    k <- fit$n_factors
    d <- k * p - k * (k - 1) / 2 + p
  } else {
    d <- p * (p + 1) / 2
  }
  data.frame(Model = label, k = ifelse(!is.null(fit$n_factors) && !is.na(fit$n_factors),
                                        fit$n_factors, "free"),
             loglik = ll, n_params = d, BIC = -2 * ll + log(n_nu) * d)
}

ic_table <- rbind(
  bic_row(fit_2f, "2-factor"), bic_row(fit_3f, "3-factor"),
  bic_row(fit_5f, "5-factor"), bic_row(fit_free, "Free")
)

knitr::kable(ic_table, digits = 1,
             caption = "Model comparison by BIC (lower is better)")
```

## Factor Loadings

```{r loadings}
# Use the 3-factor model as the primary result
fit_best <- fit_3f
Lambda <- fit_best$Lambda_hat

cat(sprintf("Factor loadings: %d lines x %d factors\n", nrow(Lambda), ncol(Lambda)))
cat("\nFirst 10 lines:\n")
print(round(head(Lambda, 10), 4))
cat("\nUnique variances (Psi, first 10):\n")
print(round(head(fit_best$Psi_hat, 10), 4))
```

# Predicting Unobserved Relationships

This is the central evaluation: how well does the factor model predict
relatedness for pairs of individuals that **never appeared together** in any
partial GRM?

## Observed vs. Unobserved Pair Classification

```{r classify-pairs}
G_truth <- G_full[demo_ids, demo_ids]
ut <- upper.tri(G_truth)
obs_mask   <- observed_mat[ut]
unobs_mask <- !obs_mask

cat(sprintf("Upper-triangle pairs: %d\n", sum(ut)))
cat(sprintf("  Observed in >= 1 GRM:  %d (%.1f%%)\n",
            sum(obs_mask), 100 * mean(obs_mask)))
cat(sprintf("  Never jointly observed: %d (%.1f%%)\n",
            sum(unobs_mask), 100 * mean(unobs_mask)))
```

## Recovery Metrics by Pair Type

We compare each model's predictions against the ground truth GRM (computed from
all 1,279 markers), separately for observed and unobserved pairs.

```{r recovery-metrics}
eval_fit <- function(fit, label) {
  Sigma <- fit$Sigma_hat[demo_ids, demo_ids]
  err_all   <- Sigma[ut] - G_truth[ut]
  err_obs   <- err_all[obs_mask]
  err_unobs <- err_all[unobs_mask]

  data.frame(
    Model  = label,
    Subset = c("Observed", "Unobserved", "All"),
    N      = c(sum(obs_mask), sum(unobs_mask), length(err_all)),
    RMSE   = c(sqrt(mean(err_obs^2)),
               sqrt(mean(err_unobs^2)),
               sqrt(mean(err_all^2))),
    Correlation = c(
      cor(Sigma[ut][obs_mask],   G_truth[ut][obs_mask]),
      cor(Sigma[ut][unobs_mask], G_truth[ut][unobs_mask]),
      cor(Sigma[ut],             G_truth[ut])
    ),
    MAE = c(mean(abs(err_obs)), mean(abs(err_unobs)), mean(abs(err_all)))
  )
}

metrics <- rbind(
  eval_fit(fit_2f,   "2-factor"),
  eval_fit(fit_3f,   "3-factor"),
  eval_fit(fit_5f,   "5-factor"),
  eval_fit(fit_free, "Free model")
)

knitr::kable(metrics, digits = 4, row.names = FALSE,
             caption = "Recovery accuracy: observed vs. unobserved pairs")
```

The key result: factor models achieve **meaningful correlation** with the ground
truth on unobserved pairs, while the free model produces **negative or near-zero
correlation** --- confirming that those entries are not identified under the
unconstrained model.

## Scatter Plots: Observed vs. Unobserved

```{r scatter-obs-unobs, fig.width=9, fig.height=5, dpi=150}
Sigma_best <- fit_best$Sigma_hat[demo_ids, demo_ids]

scatter_df <- data.frame(
  Truth     = G_truth[ut],
  Estimated = Sigma_best[ut],
  Status    = ifelse(obs_mask, "Observed", "Unobserved")
)

ggplot(scatter_df, aes(x = Truth, y = Estimated, color = Status)) +
  geom_point(alpha = 0.35, size = 1.0) +
  geom_abline(slope = 1, intercept = 0, linewidth = 0.6, linetype = "dashed") +
  scale_color_manual(values = c("Observed" = "#2166AC", "Unobserved" = "#D6604D")) +
  facet_wrap(~Status) +
  labs(
    x = "True GRM (all markers)",
    y = "Combined GRM (3-factor model)",
    title = "Predicting Pairwise Relatedness: Observed vs. Unobserved Pairs",
    subtitle = sprintf(
      "Observed: r = %.3f, RMSE = %.4f  |  Unobserved: r = %.3f, RMSE = %.4f",
      cor(scatter_df$Truth[scatter_df$Status == "Observed"],
          scatter_df$Estimated[scatter_df$Status == "Observed"]),
      sqrt(mean((scatter_df$Estimated[scatter_df$Status == "Observed"] -
                   scatter_df$Truth[scatter_df$Status == "Observed"])^2)),
      cor(scatter_df$Truth[scatter_df$Status == "Unobserved"],
          scatter_df$Estimated[scatter_df$Status == "Unobserved"]),
      sqrt(mean((scatter_df$Estimated[scatter_df$Status == "Unobserved"] -
                   scatter_df$Truth[scatter_df$Status == "Unobserved"])^2))
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 12)
  )
```

## Effect of the Number of Factors on Unobserved-Pair Prediction

```{r factor-comparison, fig.width=8, fig.height=4, dpi=150}
unobs_met <- metrics[metrics$Subset == "Unobserved" &
                       metrics$Model != "Free model", ]

if (nrow(unobs_met) > 0) {
  unobs_met$Model <- factor(unobs_met$Model,
                            levels = c("2-factor", "3-factor", "5-factor"))

  p1 <- ggplot(unobs_met, aes(x = Model, y = Correlation)) +
    geom_col(fill = "#2166AC", alpha = 0.85, width = 0.5) +
    geom_text(aes(label = round(Correlation, 3)), vjust = -0.5, size = 3.5) +
    labs(title = "Correlation with Truth\n(Unobserved Pairs)", x = NULL, y = "r") +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", size = 10))

  p2 <- ggplot(unobs_met, aes(x = Model, y = RMSE)) +
    geom_col(fill = "#D6604D", alpha = 0.85, width = 0.5) +
    geom_text(aes(label = round(RMSE, 4)), vjust = -0.5, size = 3.5) +
    labs(title = "RMSE\n(Unobserved Pairs)", x = NULL, y = "RMSE") +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", size = 10))

  gridExtra::grid.arrange(p1, p2, ncol = 2)
}
```

# Heatmap Comparison

```{r heatmaps, fig.width=9, fig.height=4, dpi=150}
n_show <- min(30, n_demo)
ids_show <- demo_ids[1:n_show]

mat_to_df <- function(mat, label) {
  sub <- mat[ids_show, ids_show]
  df <- melt(sub)
  df$Matrix <- label
  df
}

plot_df <- rbind(
  mat_to_df(G_truth,    "Ground Truth"),
  mat_to_df(Sigma_best, "CovCombR (3-factor)")
)

ggplot(plot_df, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "#F7F7F7", high = "#B2182B",
    midpoint = 0, name = "GRM"
  ) +
  facet_wrap(~Matrix) +
  coord_equal() +
  labs(x = NULL, y = NULL,
       title = "Ground Truth vs. Factor-Model Combined GRM") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text = element_blank(), axis.ticks = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid = element_blank(), legend.position = "bottom"
  )
```

# Eigenvalue Spectrum

```{r eigenvalues, fig.width=7, fig.height=4, dpi=150}
eig_truth <- eigen(G_truth,    symmetric = TRUE, only.values = TRUE)$values
eig_est   <- eigen(Sigma_best, symmetric = TRUE, only.values = TRUE)$values

n_eig <- min(20, length(eig_truth))
eig_df <- data.frame(
  Index = rep(1:n_eig, 2),
  Eigenvalue = c(eig_truth[1:n_eig], eig_est[1:n_eig]),
  Source = rep(c("Ground Truth", "3-Factor Model"), each = n_eig)
)

ggplot(eig_df, aes(x = Index, y = Eigenvalue, color = Source)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Ground Truth" = "#2166AC",
                                "3-Factor Model" = "#E74C3C")) +
  labs(x = "Eigenvalue index", y = "Eigenvalue",
       title = "Leading Eigenvalues: Ground Truth vs. Factor Model", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

# Genomic Prediction with the Combined GRM

To demonstrate downstream utility, we use the combined GRM for genomic
prediction of wheat yield (Environment 1) via ridge-regression GBLUP.

```{r genomic-prediction}
gblup <- function(y, G, prop_train = 0.7, seed = 42) {
  set.seed(seed)
  n <- length(y)
  train <- sort(sample(n, floor(n * prop_train)))
  test  <- setdiff(1:n, train)

  lambda <- 0.1
  b <- solve(G[train, train] + lambda * diag(length(train)), y[train])
  y_pred <- G[test, train, drop = FALSE] %*% b
  cor(y[test], y_pred)
}

y_demo <- Y[demo_line_idx, 1]

acc_truth <- gblup(y_demo, G_truth)
acc_3f    <- gblup(y_demo, Sigma_best)
acc_free  <- gblup(y_demo, fit_free$Sigma_hat[demo_ids, demo_ids])

pred_df <- data.frame(
  GRM = c("Ground Truth (all markers)",
          "CovCombR 3-Factor", "CovCombR Free"),
  Accuracy = round(c(acc_truth, acc_3f, acc_free), 3)
)

knitr::kable(pred_df,
             caption = "GBLUP prediction accuracy (Pearson r) for Environment 1")
```

# Discussion

## Key Findings

1. **Factor models predict unobserved relationships.** Pairs of individuals
   that never appeared together in any partial GRM still have their relatedness
   estimated with meaningful correlation to the ground truth.  The free model
   fails completely on these pairs.

2. **Low-rank structure is natural for GRMs.** Population structure in the wheat
   lines is well-captured by a few factors, making the factor model both
   statistically efficient and biologically interpretable.

3. **Downstream prediction accuracy is preserved.** Genomic prediction with the
   combined GRM approaches that of the ground-truth GRM.

## Practical Recommendations

1. **Choose $k$ based on domain knowledge** rather than `n_factors = "auto"`
   for large GRMs.  The number of major population subgroups is a good starting
   point; compare 2--3 values of $k$ by BIC.

2. **Set degrees of freedom to the number of markers** ($m - 1$) for each GRM.

3. **Use `scale_method = "estimate"`** when platforms differ in marker density.

4. **Inspect the factor loadings** ($\Lambda$) to verify that estimated factors
   correspond to known population structure.

5. **Validate with bootstrap SEs** for entries critical to downstream decisions.

# Session Information

```{r session-info}
sessionInfo()
```

# References

- Perez, P. & de los Campos, G. (2014). Genome-wide regression and prediction
  with the BGLR statistical package. *Genetics*, 198(2), 483--495.
- Rubin, D. B. & Thayer, D. T. (1982). EM algorithms for ML factor analysis.
  *Psychometrika*, 47(1), 69--76.
- VanRaden, P. M. (2008). Efficient methods to compute genomic predictions.
  *Journal of Dairy Science*, 91(11), 4414--4423.
