---
title: "Combining Incomplete Covariance Matrices: Iris Example"
author: "CovCombR Package"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Combining Incomplete Covariance Matrices: Iris Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates how to use the `CovCombR` package to combine incomplete covariance matrices using the classic iris dataset. Different studies measure different subsets of variables; CovCombR recovers the full covariance structure from these partial observations.

### The Problem

Suppose three studies each measure a different pair of iris variables:

- Study 1: Sepal.Length and Sepal.Width
- Study 2: Sepal.Length and Petal.Length
- Study 3: Sepal.Width and Petal.Width

The pairs (Sepal.Width, Petal.Length), (Petal.Length, Petal.Width), and (Sepal.Length, Petal.Width) are never jointly observed. With the **free-Σ model** those covariances would be unidentified. With the **factor model** they can be recovered through shared relationships across the overlapping variables — which is why CovCombR defaults to automatic factor-model selection.

## Example with Iris Data

### Load and Prepare

```{r load}
library(CovCombR)
data(iris)
colnames(iris) <- c("S.L", "S.W", "P.L", "P.W", "Species")
iris_virginica <- iris[iris$Species == "virginica", ]
nrow(iris_virginica)
```

### Create Incomplete Covariance Matrices

```{r create_covlist}
set.seed(1234)

CovList <- list(
  Study1 = cov(iris_virginica[sample(1:50, 20), c(1, 2)]),  # Sepal measurements
  Study2 = cov(iris_virginica[sample(1:50, 25), c(1, 3)]),  # Sepal.L + Petal.L
  Study3 = cov(iris_virginica[sample(1:50, 30), c(2, 4)])   # Sepal.W + Petal.W
)

CovList
```

The missing covariances (Sepal.Width–Petal.Length, Petal.Length–Petal.Width, Sepal.Length–Petal.Width) are never jointly observed.

### Fit the Model

The default `n_factors = "auto"` automatically selects an appropriate factor model and returns the selected fit alongside a table of all candidates.

```{r fit_model}
nu <- c(Study1 = 19, Study2 = 24, Study3 = 29)

fit <- fit_covcomb(CovList, nu = nu, se_method = "none")
fit
```

```{r show_ic_table}
fit$ic_table
```

The IC table shows how many factors were tried and why a particular model was selected. The selected model is identifiable even for the variable pairs that were never jointly observed.

### Combined Covariance Estimate

```{r extract_result}
round(fit$Sigma_hat, 4)
```

### Compare with Full Data

```{r compare_full}
Sigma_full <- cov(iris_virginica[, 1:4])

round(Sigma_full, 4)

round(fit$Sigma_hat - Sigma_full, 4)
```

### Factor Loadings (when a factor model is selected)

```{r factor_loadings}
if (!is.null(fit$Lambda_hat)) {
  round(fit$Lambda_hat, 3)
  round(fit$Psi_hat, 3)
}
```

The varimax-rotated loadings show how each variable loads onto the latent factors, and `Psi_hat` gives the residual (unique) variances.

---

## More Realistic Scenario: Different Sample Sizes

```{r different_sizes}
set.seed(2025)

CovList2 <- list(
  Small  = cov(iris_virginica[sample(1:50, 10), c(1, 2)]),
  Medium = cov(iris_virginica[sample(1:50, 25), c(2, 3)]),
  Large  = cov(iris_virginica[sample(1:50, 45), c(3, 4)])
)
nu2 <- c(Small = 9, Medium = 24, Large = 44)

fit2 <- fit_covcomb(CovList2, nu = nu2, se_method = "none")
fit2

round(fit2$Sigma_hat, 4)
```

The algorithm automatically weights studies by their degrees of freedom.

---

## Bootstrap Standard Errors

For formal inference — particularly for entries never jointly observed — bootstrap standard errors are recommended.

```{r bootstrap}
set.seed(1234)

CovList_boot <- list(
  Study1 = cov(iris_virginica[sample(1:50, 25), c(1, 2, 3)]),
  Study2 = cov(iris_virginica[sample(1:50, 30), c(2, 3, 4)]),
  Study3 = cov(iris_virginica[sample(1:50, 20), c(1, 3, 4)])
)
nu_boot <- c(Study1 = 24, Study2 = 29, Study3 = 19)

fit_boot <- fit_covcomb(
  CovList_boot,
  nu = nu_boot,
  n_factors = NULL,  # free model: bootstrap SE is implemented for the free-Sigma path
  se_method = "bootstrap",
  control = list(bootstrap = list(B = 100, seed = 1234, progress = FALSE, verbose = FALSE))
)

round(fit_boot$Sigma_se, 4)

fit_boot$bootstrap[c("B", "successes", "failures")]
```

Bootstrap standard errors account for both sampling variance and EM imputation uncertainty.

---

## Summary

The `CovCombR` workflow:

1. Provide sample covariance matrices from each study (as from `cov(X)`)
2. Specify degrees of freedom (sample size − 1) for each study
3. Call `fit_covcomb()` — the default `n_factors = "auto"` selects an identifiable model
4. Use `fit$Sigma_hat` for downstream analysis; `fit$ic_table` to review model selection
5. Add `se_method = "bootstrap"` for publication-quality uncertainty quantification

## References

- Anderson, T. W. (2003). *An Introduction to Multivariate Statistical Analysis* (3rd ed.). Wiley.
- Rubin, D. B., & Thayer, D. T. (1982). EM algorithms for ML factor analysis. *Psychometrika*, 47(1), 69–76.
- McLachlan, G. J., & Krishnan, T. (2008). *The EM Algorithm and Extensions* (2nd ed.). Wiley.

## Session Information

```{r session}
sessionInfo()
```
