---
title: "Standard Errors and Statistical Inference"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard Errors and Statistical Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
library(CovCombR)
```

## Overview

Standard errors (SEs) quantify uncertainty in your combined covariance estimates. CovCombR offers three SE methods, each with distinct trade-offs:

| Method | Speed | Reliability | Best For |
|--------|-------|-------------|----------|
| **plugin** | Fast (seconds) | Anti-conservative | Quick diagnostics |
| **bootstrap** | Slow (minutes) | Most reliable | Publication, inference |
| **sem** | Medium | Experimental | Large problems (validate!) |

**Key insight:** The default `n_factors = "auto"` selects an identifiable model before computing SEs. For variable pairs that were never jointly observed, even the bootstrap SE is valid only when the underlying covariance is identified — which requires a factor model when the free-Σ model is not identified.

---

## Which Method Should I Use?

### Decision Tree

```
Do you need standard errors at all?
├─ NO → se_method = "none" (fastest)
└─ YES
   ├─ Quick diagnostic / exploratory analysis?
   │  └─ YES → se_method = "plugin"
   │     ⚠️  WARNING: Anti-conservative for imputed entries
   │
   └─ Publication / formal inference?
      ├─ Small-medium problem (p < 50)?
      │  └─ YES → se_method = "bootstrap" (RECOMMENDED)
      │
      └─ Large problem (p > 50, bootstrap too slow)?
         └─ YES → se_method = "sem"
            ⚠️  EXPERIMENTAL: Validate against bootstrap on subset
```

---

## Method 1: Plugin Standard Errors

Plugin SEs use the asymptotic Fisher information, assuming the Wishart variance formula applies directly:

$$
\text{Var}(\hat{\Sigma}_{ij}) \approx \frac{1}{\sum_k \nu_k \cdot \mathbb{1}(i,j \in \mathcal{O}_k)}
\left( \Sigma_{ij}^2 + \Sigma_{ii}\Sigma_{jj} \right)
$$

**Critical limitation:** Plugin SEs ignore EM imputation uncertainty. They are anti-conservative for entries that rely on imputation. They return `NA` for entries never directly observed (under the free model), which is honest but not useful for factor-model fits where those entries are identified.

### Example

```{r plugin_example}
set.seed(2025)
data(iris)

# Three iris studies with partial overlap across all four variables
S_list <- list(
  Study1 = cov(iris[1:50,   1:2]),
  Study2 = cov(iris[51:100, 3:4]),
  Study3 = cov(iris[101:150, c(1, 3)])
)
nu_vec <- c(Study1 = 49, Study2 = 49, Study3 = 49)

fit_plugin <- fit_covcomb(S_list, nu = nu_vec, se_method = "plugin")

# Directly observed pair (Study 1)
fit_plugin$Sigma_hat[1, 2]
fit_plugin$Sigma_se[1, 2]

# Never jointly observed pair — factor model still estimates it, but plugin SE is NA
fit_plugin$Sigma_hat[2, 3]
fit_plugin$Sigma_se[2, 3]
```

Plugin honestly returns `NA` for pairs that have no direct coverage. Use bootstrap for valid SEs on those entries.

---

## Method 2: Bootstrap Standard Errors (Recommended)

Parametric bootstrap:

1. Simulate B datasets from the fitted model: $\mathbf{W}_k^{(b)} \sim \text{Wishart}(\nu_k, \hat{\alpha}_k \hat{\Sigma})$
2. Refit the model on each simulated dataset
3. Compute SE as the empirical standard deviation: $\text{SE}(\hat{\Sigma}_{ij}) = \text{SD}_b(\hat{\Sigma}_{ij}^{(b)})$

Bootstrap **properly accounts for** sampling variance, EM imputation uncertainty, and non-normal distributions of estimates.

### Example

```{r bootstrap_example}
set.seed(2025)

fit_boot <- fit_covcomb(
  S_list, nu_vec,
  se_method = "bootstrap",
  control = list(bootstrap = list(B = 50, seed = 2025, progress = FALSE))
)

Sigma_hat <- coef(fit_boot)
Sigma_se  <- fit_boot$Sigma_se

# Estimate and 95% CI for a never-jointly-observed pair
Sigma_hat[2, 3]
Sigma_se[2, 3]
Sigma_hat[2, 3] + c(-1, 1) * 1.96 * Sigma_se[2, 3]

# Simple Z-test
z_stat  <- Sigma_hat[2, 3] / Sigma_se[2, 3]
p_value <- 2 * pnorm(abs(z_stat), lower.tail = FALSE)
z_stat
p_value
```

### Bootstrap Diagnostics

```{r bootstrap_diagnostics}
meta <- fit_boot$bootstrap
meta[c("B", "successes", "failures")]
```

A failure rate above 10% suggests convergence issues — consider increasing `max_iter` or `ridge`.

### Configuration

```{r bootstrap_config, eval=FALSE}
fit_boot <- fit_covcomb(
  S_list, nu_vec,
  se_method = "bootstrap",
  control = list(
    bootstrap = list(
      B        = 200,    # standard for publication
      seed     = 2025,   # reproducibility
      progress = TRUE    # show progress bar
    )
  )
)
```

**Choosing B:**

- B = 100: Quick check
- B = 200: Standard (default, recommended)
- B = 500: High precision for critical applications

---

## Method 3: SEM Standard Errors (Experimental)

Supplemented EM (Meng & Rubin, 1991) uses:

$$
\mathcal{I}_{\text{obs}} = \mathcal{I}_{\text{com}} - \mathcal{I}_{\text{com}} \mathbf{R} (\mathbf{I} - \mathbf{R})^{-1}
$$

where $\mathbf{R}$ is the EM rate matrix estimated via finite differences.

### Configuration

```{r sem_config, eval=FALSE}
fit_sem <- fit_covcomb(
  S_list, nu_vec,
  se_method = "sem",
  control = list(sem = list(h = 1e-6, ridge = 1e-8))
)
```

### Validation Strategy

Always validate SEM against bootstrap on a subset before trusting it:

```{r sem_validation, eval=FALSE}
fit_sem   <- fit_covcomb(S_list, nu_vec, se_method = "sem")
fit_check <- fit_covcomb(S_list, nu_vec, se_method = "bootstrap",
                         control = list(bootstrap = list(B = 100)))

cor_check <- cor(
  as.vector(fit_sem$Sigma_se),
  as.vector(fit_check$Sigma_se),
  use = "complete.obs"
)
cor_check  # Should be > 0.95
```

```{r sem_diagnostics, eval=FALSE}
fit_sem$sem$condition_number  # Should be < 1e6
fit_sem$sem$min_eigenvalue    # Should be > 0
```

---

## Simulation: Comparing Methods

```{r comparison_setup}
set.seed(2025)
p <- 8
var_names <- paste0("Var", 1:p)

true_Sigma <- diag(p)
true_Sigma[1:4, 1:4] <- 0.6
true_Sigma[5:8, 5:8] <- 0.5
diag(true_Sigma) <- 1
dimnames(true_Sigma) <- list(var_names, var_names)

nu_vec2 <- c(Study1 = 100, Study2 = 120, Study3 = 90)

S1 <- rWishart(1, nu_vec2[1], true_Sigma[1:6, 1:6])[,,1] / nu_vec2[1]
dimnames(S1) <- list(var_names[1:6], var_names[1:6])

S2 <- rWishart(1, nu_vec2[2], true_Sigma[3:8, 3:8])[,,1] / nu_vec2[2]
dimnames(S2) <- list(var_names[3:8], var_names[3:8])

S3 <- rWishart(1, nu_vec2[3], true_Sigma[c(1,2,5,6,7,8), c(1,2,5,6,7,8)])[,,1] / nu_vec2[3]
dimnames(S3) <- list(var_names[c(1,2,5,6,7,8)], var_names[c(1,2,5,6,7,8)])

S_list2 <- list(Study1 = S1, Study2 = S2, Study3 = S3)
```

### Runtime

```{r runtime_comparison}
time_none <- system.time(
  fit_none   <- fit_covcomb(S_list2, nu_vec2, se_method = "none")
)

time_plugin <- system.time(
  fit_plugin2 <- fit_covcomb(S_list2, nu_vec2, se_method = "plugin")
)

time_boot <- system.time(
  fit_boot2 <- fit_covcomb(S_list2, nu_vec2, se_method = "bootstrap",
                           control = list(bootstrap = list(B = 50, progress = FALSE)))
)

data.frame(
  Method    = c("none", "plugin", "bootstrap (B=50)"),
  Seconds   = round(c(time_none["elapsed"], time_plugin["elapsed"], time_boot["elapsed"]), 2)
)
```

Note: B = 50 is used here for speed; use B = 200 for production.

---

## Practical Recommendations

### For Publication

Use `bootstrap` for hypothesis tests, confidence intervals, or any formal inference — especially when entries rely on imputation through the factor model.

```{r publication_example, eval=FALSE}
fit_pub <- fit_covcomb(
  S_list, nu_vec,
  se_method = "bootstrap",
  control = list(
    max_iter  = 1000,
    bootstrap = list(B = 200, seed = 12345, progress = TRUE)
  )
)
```

### For Exploration

Use `plugin` for quick diagnostics — it is instant and identifies which entries have high uncertainty.

```{r exploration_example, eval=FALSE}
fit_diag <- fit_covcomb(S_list, nu_vec, se_method = "plugin")

rel_se <- fit_diag$Sigma_se / abs(coef(fit_diag))
which(rel_se > 0.5, arr.ind = TRUE)
```

---

## Summary

| Scenario | Method | Configuration |
|----------|--------|---------------|
| Publication (small p) | bootstrap | B = 200, seed for reproducibility |
| Publication (large p) | bootstrap or sem | B = 100–200; validate SEM first |
| Exploration | plugin | Default settings |
| Model selection | none or plugin | Compare relative log-likelihoods |
| Never-observed pairs | bootstrap | Only method providing valid SEs |

**Golden Rules:**

1. Never publish with plugin SEs — anti-conservative for imputed entries
2. Bootstrap is almost always correct — worth the wait for formal inference
3. SEM requires validation — always check diagnostics and compare to bootstrap
4. Plugin NAs are informative under the free model — but factor-model fits fill them in
5. Use B = 200 as the standard bootstrap size; B = 500 for critical applications

---

## References

- Meng, X.-L., & Rubin, D. B. (1991). Using EM to obtain asymptotic variance-covariance matrices: The SEM algorithm. *Journal of the American Statistical Association*, 86(416), 899–909.
- McLachlan, G. J., & Krishnan, T. (2008). *The EM Algorithm and Extensions* (2nd ed.). Wiley.
- Efron, B., & Tibshirani, R. J. (1994). *An Introduction to the Bootstrap*. Chapman & Hall/CRC.

---

## See Also

- `vignette("CovCombR-vignette")` — Package introduction
- `vignette("advanced-configuration")` — Scale parameters and control options
- `vignette("study-design-guide")` — Designing studies for valid inference
- `?fit_covcomb` — Complete function reference
