---
title: "Designing Studies for CovCombR"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing Studies for CovCombR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
library(CovCombR)
```

## Overview

Successful use of CovCombR requires understanding *which covariances are identifiable* from your study design, and which model is appropriate to estimate them. This vignette covers:

1. **Identifiability** — when is a covariance estimable at all?
2. **Model choice** — free-Σ vs. factor model
3. **Design patterns** — recommended overlap structures
4. **Pre-flight checks** — how to validate your design before fitting

---

## Identifiability: The Core Issue

### The Free-Σ Model (Unconstrained)

In the unconstrained free-Σ model (`n_factors = NULL`), a covariance σ_ij is identified from the data *only* if variables i and j are **jointly observed in at least one study**. If no study ever sees them together, σ_ij is absent from the observed-data likelihood; the EM will return a number, but that number depends entirely on initialization and should not be trusted.

**Connectivity is necessary but not sufficient.** Even when the observation graph is connected (every variable is reachable from every other), some covariances can be unidentified under the free model if the connecting path never includes the pair jointly.

### The Factor Model (Default)

Under a k-factor model (Σ = ΛΛᵀ + Ψ), identifiability is extended. If variable i is observed alongside variables {a, b, …} in one study, and variable j is observed alongside those same variables in another study, their shared factor loadings are identified through the overlap — and σ_ij = Λ[i,·]·Λ[j,·] becomes identified even though (i, j) are never jointly observed.

**Example — the chain design:**

```
Study 1: {V1, V2}     → identifies λ₁, λ₂ relative to each other
Study 2: {V2, V3}     → identifies λ₂, λ₃ relative to each other
Shared anchor (V2)    → identifies σ₁₃ = λ₁λ₃ via λ₂
```

Under a 1-factor model, σ₁₃ = σ₁₂ · σ₂₃ / σ₂₂ is identified from the two studies even though V1 and V3 are never observed together.

### Default Behavior

`fit_covcomb()` defaults to `n_factors = "auto"`, which fits factor models of increasing complexity and selects the best by BIC. This means:

- If your data support the free model (full pairwise coverage), it will typically be selected.
- If your data have gaps (chain patterns, sparse designs), a factor model will be selected, giving valid estimates where the free model would not.

---

## Minimum Requirements

### Hard Constraints

| Requirement | Minimum | Why |
|-------------|---------|-----|
| **Studies (K)** | ≥ 2 | Cannot combine a single study |
| **Variables (p)** | ≥ 3 | Covariance needs at least 3 variables |
| **Sample size (ν_k)** | ≥ p_k | Wishart requires ν ≥ dimension |
| **Graph connectivity** | All connected | Each variable reachable from every other |

Graph connectivity is required for both the free model and the factor model. Variables in disconnected components cannot share factor structure and remain unidentifiable.

### Practical Guidelines

| Guideline | Recommended | Impact |
|-----------|-------------|--------|
| **Studies (K)** | ≥ 3 | Better factor identification, stability |
| **Overlap per study** | > 40% | Reduces imputation uncertainty |
| **Sample size (ν_k)** | ≥ 10 × p_k | Adequate precision per study |

---

## Design Patterns

### Pattern 1: Complete Overlap (Ideal)

All studies observe all variables. No imputation needed; the free model is always valid.

```
Study 1: {V1, V2, V3, V4, V5}
Study 2: {V1, V2, V3, V4, V5}
Study 3: {V1, V2, V3, V4, V5}
```

Use when: simulation studies, method validation, or when data collection is fully controlled.

### Pattern 2: Balanced Overlap (Recommended)

Each study observes ~60–80% of variables with systematic rotation; all pairs observed in at least one study.

```
Study 1: {V1, V2, V3, V4, V5, V6, V7, V8}
Study 2: {V3, V4, V5, V6, V7, V8, V9, V10}
Study 3: {V1, V2, V5, V6, V7, V8, V9, V10}
```

With full pairwise coverage, the free model is valid and auto-selection will likely favor it. Use for multi-site studies or meta-analysis.

### Pattern 3: Hierarchical Blocks

Studies organized in blocks with a bridge study connecting them.

```
Study 1 (Block A): {V1, V2, V3, V4}
Study 2 (Block B): {V5, V6, V7, V8}
Study 3 (Bridge):  {V2, V3, V6, V7}   ← connects both blocks
```

The bridge study ensures connectivity. Cross-block pairs (e.g., V1–V5) are not directly observed, so the factor model will typically be selected and will estimate them via shared factor structure.

### Pattern 4: Star Topology

One large hub study, several focused spoke studies.

```
Hub:     {V1, V2, V3, V4, V5, V6, V7, V8}
Spoke 1: {V1, V2, V3}
Spoke 2: {V4, V5, V6}
Spoke 3: {V7, V8}
```

The hub provides all pairwise coverage; spokes increase precision for specific subsets. Hub quality is critical.

### Pattern 5: Chain Topology — Now Supported via Factor Model

Previously treated as problematic, chain designs are now supported when using the factor model.

```
Study 1: {V1, V2, V3}
Study 2: {V2, V3, V4}
Study 3: {V4, V5, V6}
```

Under the free model, pairs like V1–V6 are unidentified. Under a factor model, the shared variables (V2, V3, V4) serve as anchors that propagate factor loadings across the chain. The default `n_factors = "auto"` will select a factor model automatically when needed.

```{r chain_comparison}
set.seed(2025)
var_names <- c("V1", "V2", "V3")

# True 1-factor covariance: lambda = c(0.8, 0.9, 0.7)
lam <- c(0.8, 0.9, 0.7)
true_Sigma <- outer(lam, lam) + diag(c(0.36, 0.19, 0.51))
dimnames(true_Sigma) <- list(var_names, var_names)

# Chain pattern: V1-V2 in Study 1, V2-V3 in Study 2
S1 <- rWishart(1, 200, true_Sigma[1:2, 1:2])[,,1] / 200
S2 <- rWishart(1, 200, true_Sigma[2:3, 2:3])[,,1] / 200
dimnames(S1) <- list(var_names[1:2], var_names[1:2])
dimnames(S2) <- list(var_names[2:3], var_names[2:3])

S_list_chain <- list(S1 = S1, S2 = S2)
nu_chain     <- c(S1 = 200, S2 = 200)

# Free model: V1-V3 is NOT identified
fit_free <- fit_covcomb(S_list_chain, nu_chain, n_factors = NULL, se_method = "none")

# Factor model (k=1): V1-V3 IS identified
fit_fa <- fit_covcomb(S_list_chain, nu_chain, n_factors = 1, se_method = "none")

# Auto-selection picks the factor model
fit_auto <- fit_covcomb(S_list_chain, nu_chain, se_method = "none")

data.frame(
  Model       = c("Free Sigma", "1-factor", "Auto"),
  sigma_V1_V3 = round(c(
    fit_free$Sigma_hat["V1", "V3"],
    fit_fa$Sigma_hat["V1", "V3"],
    fit_auto$Sigma_hat["V1", "V3"]
  ), 4),
  True_sigma_V1_V3 = round(true_Sigma["V1", "V3"], 4)
)
```

The free model returns an arbitrary value for V1–V3 (depends on initialization). The factor model recovers an estimate close to the true value; auto-selection finds the same solution.

---

## Identifiability Diagnostics

### Connectivity Check

CovCombR checks graph connectivity automatically and warns if disconnected. You can also check manually:

```{r connectivity_check}
check_connectivity <- function(S_list) {
  all_vars <- sort(unique(unlist(lapply(S_list, rownames))))
  p        <- length(all_vars)
  var_map  <- setNames(seq_len(p), all_vars)
  adj      <- matrix(0L, p, p)

  for (S in S_list) {
    idx <- var_map[rownames(S)]
    adj[idx, idx] <- 1L
  }
  diag(adj) <- 0L

  visited <- logical(p)
  n_comp  <- 0L
  for (start in seq_len(p)) {
    if (visited[start]) next
    n_comp <- n_comp + 1L
    queue  <- start
    visited[start] <- TRUE
    while (length(queue) > 0) {
      cur      <- queue[1]; queue <- queue[-1]
      nbrs     <- which(adj[cur, ] == 1L & !visited)
      visited[nbrs] <- TRUE
      queue    <- c(queue, nbrs)
    }
  }
  list(is_connected = n_comp == 1L, num_components = n_comp, variables = all_vars)
}

# Example: balanced overlap design
S1 <- matrix(1, 5, 5); rownames(S1) <- colnames(S1) <- paste0("V", 1:5)
S2 <- matrix(1, 5, 5); rownames(S2) <- colnames(S2) <- paste0("V", 3:7)
S3 <- matrix(1, 4, 4); rownames(S3) <- colnames(S3) <- paste0("V", c(1, 2, 6, 7))

check_connectivity(list(S1, S2, S3))
```

### Pairwise Coverage Matrix

```{r coverage_matrix}
compute_coverage <- function(S_list, nu) {
  all_vars <- sort(unique(unlist(lapply(S_list, rownames))))
  p        <- length(all_vars)
  var_map  <- setNames(seq_len(p), all_vars)
  coverage <- matrix(0, p, p, dimnames = list(all_vars, all_vars))

  for (nm in names(S_list)) {
    idx  <- var_map[rownames(S_list[[nm]])]
    coverage[idx, idx] <- coverage[idx, idx] + nu[nm]
  }
  diag(coverage) <- NA
  coverage
}

nu_ex   <- c(S1 = 100, S2 = 120, S3 = 90)
cov_mat <- compute_coverage(list(S1 = S1, S2 = S2, S3 = S3), nu_ex)

cov_mat

n_pairs    <- sum(!is.na(cov_mat)) / 2
n_observed <- sum(cov_mat > 0, na.rm = TRUE) / 2
sprintf("Directly observed pairs: %.1f%%", 100 * n_observed / n_pairs)
```

Pairs with zero coverage are not directly identified under the free model. The factor model can still estimate them if the graph is connected and the factor structure provides a path.

---

## Pre-Flight Checklist

Before calling `fit_covcomb()`:

```{r checklist, eval=FALSE}
# 1. Check connectivity
conn <- check_connectivity(S_list)
stopifnot(conn$is_connected)

# 2. Check coverage
cov_mat <- compute_coverage(S_list, nu)
pct <- 100 * sum(cov_mat > 0, na.rm = TRUE) / sum(!is.na(cov_mat))
sprintf("Coverage: %.1f%%", pct)

# 3. Fit with auto-selection and inspect IC table
fit <- fit_covcomb(S_list, nu, se_method = "none")
fit$ic_table

# 4. Check condition number of result
kappa(fit$Sigma_hat)
# > 1e6 → ill-conditioned; may need more overlap or larger nu
```

---

## Power Considerations

**What affects estimation quality?**

1. **Sample size (ν):** SE ∝ 1/√(Σν_k) for directly observed pairs; factor model SEs scale similarly.

2. **Number of studies (K):** More studies increase both coverage and factor loading precision. Benefits diminish after K ≈ 5.

3. **Overlap percentage:** More overlap → less imputation → smaller SEs for all methods.

4. **Factor dimensionality:** If the true model has k* factors, fitting k < k* causes bias; fitting k > k* adds unnecessary parameters and inflates variance. Auto-selection by BIC guards against both errors.

**Rule of thumb:** Aim for each variable to appear in at least 3 studies, with ≥ 40% overlap per study pair.

---

## Recommendations Summary

**Use `n_factors = "auto"` (default) unless you have a specific reason to override.**

When the data support the free model (all pairs observed), auto-selection typically finds it. When the data have gaps (chain patterns, block designs with bridges), auto-selection fits an appropriate factor model.

Override the default only when:

- You have strong domain knowledge about the factor dimensionality → use `n_factors = k`
- You have confirmed full pairwise coverage and want the exact free-Σ MLE → use `n_factors = NULL`
- You want AIC instead of BIC for model selection → use `control = list(ic = "AIC")`

---

## See Also

- `vignette("CovCombR-vignette")` — Package introduction with worked example
- `vignette("standard-errors")` — SE method selection
- `vignette("statistical-methods")` — Full mathematical details including factor model
- `?fit_covcomb` — Complete function reference
